\texttt{davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2022-10-16 paper/main.tex
add package structure figure
updates to implementation details section

\documentclass[preprint,12pt, a4paper]{elsarticle}

@@

\subsection{Software architecture}
The \texttt{davos} package consists of two interdependent subpackages. The first, \texttt{davos.core}, comprises a set of modules that implement\comment{provide?} the bulk of the package's core functionality, including pipelines for installing and validating packages, custom parsers for the \texttt{smuggle} statement (see Section \ref{subsec:smuggle}) and onion comment (see Section \ref{subsec:onion}), and a runtime interface for configuring \texttt{davos}'s behavior (see Section \ref{subsec:config}). However, certain critical\comment{other important} aspects of this functionality require (often substantially) different implementation approaches depending on various properties of the notebook environment in which \texttt{davos} is used (e.g., whether the frontend is provided by Jupyter or Google Colaboratory, or which version of IPython \cite{PereGran07} is used by the notebook kernel). To deal with this, environment-dependent parts of core features and behaviors are isolated and abstracted to ``helper functions'' in the \texttt{davos.implementations} subpackage. This second subpackage defines multiple, interchangeable versions of each helper function, organized into\comment{per-environment} modules by the conditions that trigger their use. At runtime, \texttt{davos} detects various features in the notebook environment and selectively imports a single version of each helper function into the top-level \texttt{davos.implementations} namespace, allowing \texttt{davos.core} modules to access the correct implementations for the current notebook environment from a single, constant location. An additional benefit of this design pattern is that it allows maintainers or contributors to extend \texttt{davos} to support new, updated, or custom notebook variants simply by creating a new \texttt{davos.implementations} module with any necessary tweaks to existing helper functions.

@@

\comment{
- js_functions.py?
- how parser is registered and deregistered?
- stub files?
- test suite?
- packaged with new PEP \_\_\_ standard (pyproject.toml + setup.cfg; no setup.py)?
}
% Give a short overview of the overall software architecture; provide a pictorial component overview or similar (if possible). If necessary provide implementation details.

@@

\subsection{Implementation details}\label{subsec:implementation}
Functionally, importing \texttt{davos} appears to define ``\texttt{smuggle}'' as a Python keyword, similar to ``\texttt{import}'', ``\texttt{def}'', or ``\texttt{return}''. It also appears to cause comments to be parsed and potentially able to affect the code's behavior, which they normally are not. However, \texttt{davos} doesn't actually modify the rules of Python's parser or lexical analyzer---in fact, modifying the Python grammar isn't possible at runtime, as doing so would require rebuilding the interpreter. Instead, \texttt{davos} leverages the IPython notebook backend to implement the \texttt{smuggle} statement and onion comment via a combination of namespace injections and its own (far simpler) custom parser.

@@

The \texttt{smuggle} keyword can be enabled and disabled at any time by ``activating'' and ``deactivating'' \texttt{davos} (see Section \ref{subsec:config}, above); when \texttt{davos} is first imported, it is activated automatically. Activating \texttt{davos} triggers two things: (1) the \texttt{smuggle()} function is injected into the IPython user namespace, and (2) the \texttt{davos} parser is registered as a custom IPython input transformer. IPython preprocesses all executed code as plain text before it is sent to the Python parser, in order to handle special constructs like \texttt{\%magic} and \texttt{!shell} commands. \texttt{davos} hooks into this process to transform \texttt{smuggle} statements into syntactically valid Python code. The \texttt{davos} parser uses a complex regular expression \cite{Thom68} to match lines of code containing \texttt{smuggle} statements (and, optionally, onion comments), extract relevant information from their text, and replace them with equivalent calls to the \texttt{smuggle()} function. For example, if the code visible to the user reads 
\begin{verbatim}
smuggle numpy as np    # pip: numpy>1.16,<=1.20 -vv,
\end{verbatim} 
the code that is actually executed by the Python interpreter would be 
\begin{verbatim}
smuggle(name="numpy", as_="np", installer="pip", args_str="""numpy>1.16,<=1.20 -vv""", installer_kwargs={`editable': False, `spec': `numpy>1.16,<=1.20', `verbosity': 2}).
\end{verbatim}

>>>>

\documentclass[preprint,12pt,a4paper]{elsarticle}

@@

\subsection{Software architecture}
% Give a short overview of the overall software architecture; provide a pictorial component overview or similar (if possible). If necessary provide implementation details.
The \texttt{davos} package consists of two interdependent subpackages. The first, \texttt{davos.core}, comprises a set of modules that implement\comment{provide?} the bulk of the package's core functionality, including pipelines for installing and validating packages, custom parsers for the \texttt{smuggle} statement (see Section \ref{subsec:smuggle}) and onion comment (see Section \ref{subsec:onion}), and a runtime interface for configuring \texttt{davos}'s behavior (see Section \ref{subsec:config}). However, certain critical\comment{other important} aspects of this functionality require (often substantially) different implementation approaches depending on various properties of the notebook environment in which \texttt{davos} is used (e.g., whether the frontend is provided by Jupyter or Google Colaboratory, or which version of IPython \cite{PereGran07} is used by the notebook kernel). To deal with this, environment-dependent parts of core features and behaviors are isolated and abstracted to ``helper functions'' in the \texttt{davos.implementations} subpackage. This second subpackage defines multiple, interchangeable versions of each helper function, organized into\comment{per-environment} modules by the conditions that trigger their use. At runtime, \texttt{davos} detects various features in the notebook environment and selectively imports a single version of each helper function into the top-level \texttt{davos.implementations} namespace, allowing \texttt{davos.core} modules to access the correct implementations for the current notebook environment from a single, constant location. An additional benefit of this design pattern is that it allows maintainers or contributors to extend \texttt{davos} to support new, updated, or custom notebook variants simply by creating a new \texttt{davos.implementations} module with any necessary tweaks to existing helper functions.

@@

\comment{
- js_functions.py?
- how parser is registered and deregistered?
- stub files?
- test suite?
- packaged with new PEP \_\_\_ standard (pyproject.toml + setup.cfg; no setup.py)?
}

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/package_structure}
\caption{\small \textbf{Package structure.}}
\label{fig:package-structure}
\end{figure}

@@

\subsection{Implementation details}\label{subsec:implementation}
Functionally, importing \texttt{davos} appears to define ``\texttt{smuggle}'' as a Python keyword, similar to ``\texttt{import}'', ``\texttt{def}'', or ``\texttt{return}''. It also appears to cause comments to be parsed, and their contents potentially able to affect code behavior, which they normally are not. However, \texttt{davos} doesn't actually modify the rules of Python's parser or lexical analyzer---in fact, modifying the Python grammar isn't possible at runtime, as doing so would require rebuilding the interpreter. Instead, \texttt{davos} leverages the IPython notebook backend to implement the \texttt{smuggle} statement and onion comment via a combination of namespace injections and its own (far simpler) custom parser.

@@

The \texttt{smuggle} keyword can be enabled and disabled at any time by ``activating'' and ``deactivating'' \texttt{davos} (see Section \ref{subsec:config}, above). When \texttt{davos} is first imported, it is activated automatically. Activating \texttt{davos} triggers two things: (1) the \texttt{smuggle()} function is injected into the IPython user namespace, and (2) the \texttt{davos} parser is registered as a custom IPython input transformer. IPython preprocesses all executed code as plain text before it is sent to the Python parser, in order to handle special constructs like \texttt{\%magic} and \texttt{!shell} commands. \texttt{davos} hooks into this process to transform \texttt{smuggle} statements into syntactically valid Python code. The \texttt{davos} parser uses a complex regular expression \cite{Thom68} to match lines of code containing \texttt{smuggle} statements (and, optionally, onion comments), extract relevant information from their text, and replace them with equivalent calls to the \texttt{smuggle()} function. For example, if a user runs a notebook cell containing
\begin{verbatim}
smuggle numpy as np    # pip: numpy>1.16,<=1.20 -vv,
\end{verbatim}
the code that is actually executed by the Python interpreter would be
\begin{verbatim}
smuggle(name="numpy", as_="np", installer="pip", args_str="""numpy>1.16,<=1.20 -vv""", installer_kwargs={`editable': False, `spec': `numpy>1.16,<=1.20', `verbosity': 2}).
\end{verbatim}
Because the \texttt{smuggle()} function is defined in the notebook namespace in order for this to work, it is also possible (though never necessary) to call it directly. Deactivating \texttt{davos} will delete the name ``\texttt{smuggle}'' from the namespace, unless it has been overwritten and no longer refers to the \texttt{smuggle()} function, as well as deregister the \texttt{davos} parser from the set of input transformers that are run when each notebook cell is executed. While the overhead added by the \texttt{davos} parser is de minimis, this may be useful, for example, when optimizing or precisely profiling code.
